Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-08-13 thesis/introduction.tex
thesis updates and ludlow notebook

Dwarf galaxies host, in many ways, the most extreme galactic
environments in the universe. These little galaxies are typically
defined to fainter than the LMC or SMC \citep[\(M_V \gtrsim -18\) ,
e.g.,][]{hodge1971, mcconnachie2012}. As the smallest class of galaxies,
dwarfs are the most numerous galaxies. Dwarf galaxies are highly
\emph{dark-matter dominated}, with mass to light ratios sometimes
exceeding 100 to 1000 \(M_\odot/ L_\odot\). Because of their small total
masses, many dwarf galaxies are \emph{quenched}, with little to no
recent star formation. Their stellar populations are \emph{relics} from
the early universe, consisting of many of the oldest and most metal poor
stars. Understanding the properties of dwarf galaxies thus has
implications across astronomy, from cosmological structure formation on
the smallest scale to extremely metal-poor stellar populations to
chemical evolution to galactic dynamics. Of the classical dwarfs, the
Sculptor and Ursa Minor dwarf spheroidal galaxies stand out with a more
extended density profile relative to an exponential, hinting at tidal
effects. As a case study in tidal effects and our understanding of dwarf
galaxy formation and evolution, we aim to understand the dynamical
history of these galaxies.

@@

We only understand a tiny fraction of the composition of the universe.
The leading theory of cosmology, \(\Lambda\)CDM (cold dark matter),
posits that the universe is composed of about 68\% dark energy
(\(\Lambda\)), 27\% dark matter (DM), and 5\% regular baryons\footnote{Astronomers
  like to change definitions of words. \emph{Baryons} here means
  baryons+leptons, i.e.~any standard model massive fermion. The photon
  energy density is negligible today} \citep{planckcollaboration+2020}.
While the composition of dark matter and dark energy remains elusive, we
know their general properties. Dark energy causes the acceleration of
the expansion of the universe on large scales. We do not discuss dark
energy here---it does not substantially affect the local group today.
Dark matter, instead, makes up the vast majority of mass in galaxies.
Typically, galaxies have baryonic to dark matter ratios of between 1:5
to beyond 1:1000 for faint dwarf galaxies. In \(\Lambda\)CMD, dark
matter is assumed to only interact gravitationally, passing through
matter without effect (transparent to light, or \emph{dark}). Dark
matter is also \emph{cold}, i.e.~typical velocities much smaller than
the speed of light in the early universe. Implications of dark matter
properties range from cosmological structural formation, galaxy
structure, and galaxy interactions.

@@

Dark matter halos, and their associated galaxies, rarely evolve in
isolation. Instead, structure formation is \emph{hierarchical}. Small
dark matter halos collapse first and hierarchically merge into
progressively larger halos
\citep[e.g.,][]{blumenthal+1984, blumenthal+1986, white+rees1978, white+frenk1991, somerville+dave2015}.
Structure formation happens on a wide range of scales. The largest
structures become the cosmic web---composed of voids, filaments, and
clusters---and the smallest structures directly observable host dwarf
galaxies. Hierarchical assembly is evident through the large scale
structure of the universe, remnants of past mergers within the milky
way, and tidal disruption of dwarf galaxies and their streams around
nearby galaxies.

Small-scale structure formation is sensitive to deviations from
\(\Lambda\)CDM cosmology \citep[e.g.,][]{bechtol+2022}. Many alternative
models, such as warm dark matter or self-interacting dark matter, may
smooth out small-scale features and reduce the abundance of small halos
or change their structure \citep[e.g.,][]{lovell+2014}. Dwarf galaxies,
occupying the smallest dark matter halos, are promising windows into
small-scale cosmological features. Alternative dark matter properties
can influence dwarf galaxy structure, formation, and tidal evolution.
The nearby dwarf galaxies of the local group, with detailed
observations, present a promising opportunity to understand the
evolution of these objects and test our understanding of cosmology. To
understand the evolution of these objects, we need to understand the
general predictions from \(\Lambda\)CDM for the properties of dwarf
galaxies.

@@

The total mass of an NFW profile diverges, so halos are commonly
characterized by an overdensity criterion. The virial mass, \(M_{200}\),
is defined as the mass within a region \(r_{200}\) containing a mean
enclosed density 200 times\footnote{200 smells like an arbitrary value.
  From approximate analytic arguments about the mean density of a
  self-gravitating spherical collapse halo in an expanding universe, the
  expected virial overdensity is XXX, rounding up to 200 (REFS).} the
critical density of the universe: \begin{equation}{
M_{200} =(4\pi/3) \ r_{200}^3\ 200\rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z) / 8\pi G
}\end{equation} A standard second parameter is the halo concentration,
\(c=r_{200} / r_s\) describing how the characteristic size of the halo
compares to its virial radius. In this case, the scale density is a
function of \(c\) alone,
\(\rho_s = (200/3)\,\rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\)
\citep{NFW1996}. However, we characterize halos by their circular
velocity profiles. The circular velocity,
\(v_{\rm circ}(r) = \sqrt{G M(r) / r}\), reaches a maximum of
\(v_{\rm max}\) at radius \(r_{\rm max} = \alpha\,r_s\) where
\(\alpha\approx2.16258\). \(v_{\rm max}\) is related to both the total
halo mass and the observed line of sight velocity disperion, and
\(r_{\rm max}\) relates to the scale radius.

While an NFW profile has two free parameters (\(M_{200}\) and \(c\)),
these are not independent. Smaller dark matter halos often collapse
earlier, when the universe was denser. As a result, small subhalos tend
to be more concentrated \citep[e.g.,][]{NFW1997}. The relationship
between \(M_{200}\) and c, or the mass-concentration relation, how more
massive halos become less concentrated in a predictable way
\citep[e.g.,][]{bullock+2001, ludlow+2016}. While there is a general
trend in concentration with mass, the concentration values still have
substantial scatter. Other parameters such as the halo spin or shape or
slight changes in the halo density profile add additional variability to
the present-day distribution of halos \citep[see
e.g.,][darkEXP]{navarro+2010, dicintio+2013, dutton+maccio2014}.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/example_density_profiles.pdf}}
\caption[Example density profiles]{Example stellar and dark matter
density profiles for a Sculptor-like galaxy. The dark matter is more
extended and massive than the star across the entire
galaxy.}\label{fig:nfw_density}
\end{figure}

@@

Both cosmology and observations find fundamental relationships between
properties of galaxies, particularly for the total mass and luminosity.
A key prediction of \(\Lambda\)CDM is the stellar mass halo mass
relation, describing the amount of stellar mass forms in a (sub)halo of
a given size. In particular, the SMHM relation grows especially steep in
the dwarf galaxy regime---many dwarf galaxies are formed in halos of
similar masses. Fig.~\ref{fig:smhm} shows the stellar mass versus
\(v_{\rm max}\) maximum circular velocity (proxy for halo mass) for
apostle dwarfs from \citet{fattahi+2018}. While there is some scatter,
the range of possible \(v_{\rm max}\) is fairly narrow across \(\sim 5\)
decades in stellar mass. Thus, if we know the initial stellar mass of a
dwarf galaxy, we also know, at some level, its dark matter mass.

Several challenges complicate a simple SMHM trend including environment,
assembly history, and tidal effects. By being closer to a massive host,
most dwarfs quench earlier, resulting in lower stellar masses
\citep[e.g.,][]{christensen+2024}. Additionally, the time of formation
(relative to reionization) can influence the resulting stellar content
\citep{kim+2024}. Finally, tides influence both the dark matter and
stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
Consequently, tides may reduce the halo mass more than the stellar mass,
adding additional scatter to the SMHM trend, particularly for Milky Way
satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
tides on local group dwarf galaxies is essential to determining where
and how these galaxies formed in a cosmological context.

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715100059332.png}
\caption[Stellar-mass halo-mass relation]{The stellar mass halo mass
relation, where \(v_{\rm max}\) is related to the halo mass. Right: the
mass-concentration relation for NFW halos, but parameterized in terms of
\(r_{\rm max}\) and \(v_{\rm max}\). Together, these plots represent the
cosmologically expected properties of underlying dark matter halos in
any dwarf galaxy. The velocity dispersion directly constrains the mass
contained within a half-light radius, so the underlying halo in
\(\Lambda\)CDM is well-constrained. Adapted from figure 1 of
\citet{fattahi+2018}, stellar mass and \(v_{\rm max}\) of APOSTLE
satellites at peak \(v_{\rm max}\).}\label{fig:smhm}
\end{figure}

@@

Dwarf galaxies have long challenged questioned theories of galaxy
formation. The first discovered dwarf galaxies in 1938\footnote{technically,
  the LMC and SMC may be classified as dwarf galaxies, but these were
  likely always known to humans at southern latitudes.}, Sculptor and
Fornax, already presented such a puzzle. Having no close analogues,
\citet{shapley1938} present Sculptor and Fornax as a new type of
\emph{stellar system} resembling the Magellanic clouds and globular
clusters, but not attempting to speculate on the exact nature. While
generally understood to be galaxies based on the inferred luminosities
and sizes, the nature and formation remained unclear
\citep[e.g.,][]{hodge1971}.

The earliest spectroscopic work hinted that dwarf galaxies may contain
substantial dark matter. From the first measurements of velocity
dispersions
\citep[e.g.,][]{aaronson1983, gallagher+wyse1994, pryor1996}, the
inferred mass-to-light ratios were about 10 times the expectation for
globular clusters \citep{matteo1998}. Subsequently, many theories
attempting to understand the formation and observed properties of these
objects were proposed. Examples included that dwarf galaxies were formed
tidally in the gas clouds of satellites; dwarf galaxies are undergoing
tidal dissolution resulting in extreme mass-to-light ratios
\citep[e.g.,][]{pryor1996}, or Since then, we have known that dwarf
galaxies are among the darkest objects in the universe, and
understanding their properties is critical to understanding the
universe.

Dwarf galaxies span a large range of physical sizes, luminosities, and
morphologies. Broadly, there are three classes of dwarf galaxies based
on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
\textbf{bright dwarfs galaxies} with magnitudes
\(-18 \lesssim M_V \lesssim  -14\), often exhibit irregular morphologies
and recent star formation. Fig.~\ref{fig:galaxy_images} shows the LMC as
an example of an irregular (and slightly bright) dwarf galaxy displaying
a bar. \textbf{Classical dwarf} galaxies occupy intermediate
luminosities ( \(-14 \lesssim M_V  \lesssim -7.7\)). Typically, these
systems are old, gas-poor, and spheroidal. All Milky Way satellites
discovered before digital sky surveys are classicals. The 12 classical
dwarfs satellites of our galaxy are Sagittarius, Fornax, Leo I,
Sculptor, Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici
I, Sextans I, and Crater II.\footnote{While formally the dwarf galaxy
  names we discuss contain'' dwarf spheroidal'' (dSph), e.g.~Sculptor
  dSph, we omit this suffix for brevity.} The \textbf{ultra-faint}s
occupy the very faintest magnitudes (\(-7.7 \lesssim M_V\)). These
galaxies have minuscule stellar masses, tend to be more compact, and are
the darkest known galaxies. An example is Boötes V as shown in
Fig.~\ref{fig:galaxy_images}. Altogether, dwarf galaxies span about 7
orders of magnitude in stellar mass.

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.pdf}
\caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Sculptor (DES
2), Ursa Minor (UNWISE with Gaia point sources over-plotted), and Bootes
V (SDSS). While the LMC is very prominent in the sky, even the classical
dwarf galaxies are not obvious except in terms of star counts
(e.g.,Fig.~\ref{fig:scl_selection}). All (non-LMC) dwarfs occupy the
central third to sixth of the image.}\label{fig:galaxy_images}
\end{figure}

@@

\emph{Gaia} has redefined astrometry, providing photometry, positions,
proper motions, and parallaxes for over 1 billion stars
\citep{gaialcollaboration+2021}. Launched in 2013, \emph{Gaia} is a
space-based, all-sky survey telescope situated at the Sun-Earth L2
Lagrange point \citep{gaiacollaboration+2016}. While \emph{Gaia}
completed its space-based mission in 2025, two forthcoming data releases
remain. Determining absolute parallax measurement is facilitated by the
observation that stars in different regions of the sky are affected by
parallax motion with different phases. By imagining two regions
separated by 106.5 degrees on the same focal plane, \emph{Gaia} measures
tiny changes in relative positions of stars across small and large
angles. Combining measurements from multiple epochs across several
years, an absolute all-sky reference frame is derived from which
parallax and proper motions are derived \citep{gaiacollaboration+2016}.
In addition to astrometry, \emph{Gaia} measures photometry in the wide
\emph{G} band (330-1050nm) and colours from the blue photometer (BP,
330-680 nm) and red photometer (RP, 640-1050 nm). \emph{Gaia}
additionally provides low resolution BP-RP spectra and radial velocity
measurements of bright stars (\emph{Gaia} radial velocity magnitudes
\textless16) \citep{gaiacollaboration+2016}. For this work,
\emph{Gaia}'s most relevant measurement are \(G\) magnitude,
\(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{\(\mu_\alpha\)
  and \(\mu_\delta\) are the apparent changes in \(\alpha\) and
  \(\delta\). \(\mu_{\alpha*}\) allows where
  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
  effects in \(\alpha\).}

\emph{Gaia} has revolutionized many fields of astronomy including the
Milky Way and the Local Group. For example, 6D kinematic classification
of Milky Way stars has lead to the discovery of past mergers or Milky
Way building blocks like \emph{Gaia}-sausage Enceladus
\citep[e.g.,][]{helmi+2018}, out-of-equilibrium structures like the
\emph{Gaia} snail {[}e.g., \citet{antoja+2018l}, and effects of the
Milky Way's bulge and bar \citep{hunt+vasiliev2025}. Moving to the Milky
Way halo, numerous streams \citep{bonaca+price-whelan2025}. \emph{Gaia}
has revealed the hierarchical formation and complex, evolving structure
of our own galaxy.

For Milky Way satellites, \emph{Gaia} has enabled well-constrained
orbital analysis and facilitated precise stellar membership
determination. Before, proper motions of dwarf galaxies were spase or
undetermined and measured by the Hubble Space Telescope or other methods
for only a few systems. \emph{Gaia} allowed for the first systematic
determinations of Milky Way satellite proper motions, and of
unprecedented precision \citep{MV2020a, pace+li2019}. While the proper
motion uncertainty on a typical dwarf member star is often with large
uncertainties, by combining the proper motions for 100s to 1000s of
stars in \emph{Gaia}, incredibly precise proper motion measurements can
be determined, sometimes only limited by \emph{Gaia}'s systematic error
floor \citep[e.g.,][]{MV2020a}. Proper motions have ushered in a new
dynamical era for MW satellites studies, where we often have precise
determinations of the dwarf galaxy's recent orbit under a given MW
potential. In addition, \emph{Gaia} helps seperate out contaiminating MW
foreground stars. By measuring parallaxes or proper motions
significantly different from the dwarf galaxy, many stars can be
immediately removed as non-members
\citep[e.g.,][\citet{jensen2024}]{battaglia+2022}.

@@

Of local dwarfs, the classical systems still remain the best studied and
with most strongly constrained parameters. While extending 1-2 degrees
across the sky, making deriving properties trickier
\citep[e.g.,][]{mateo1998}, these dwarfs contain large numbers of bright
(giant) stars, allowing thousands of stars to be observed with deep
photometry and spectroscopy \citep[e.g.,][]{tolstoy+2023, pace+2020}. As
a result, the determination of fundamental properties such as the
position, size, orientation, distances, proper motions, line-of-sight
(LOS) velocities, and velocity dispersions \(\sigma_v\), are all
relatively well constrained today and show excellent convergence among
different works. However, ongoing research continues to redefine our
understanding of the detailed structure of Milky Way satellites, hinting
that these objects may be more complex than at first glance.

Observations of dwarf galaxies have been the origin of several
\emph{small-scale} problems in cosmology
\citep[e.g.,][]{bullock+boylan-kolchin2017}. For example, the mismatch
between the expected number of dwarf galaxies from simulations and the
observed abundance was known as the \emph{missing satellites problem}.

\emph{Core Cusp Problem}. One of the open questions in dwarf galaxy
evolution is the formation of dark matter \emph{cores} where the density
becomes constant instead \(\rho \sim 1\). A leading theory is cores form
through of baryonic effects and feedback. If dark matter deviates from
\(\Lambda\)CDM, then the NFW density profile may not describe halos
anymore. \emph{Warm} dark matter, relativistic in the early universe but
cooler now, smooths out small-scale features and softens the cusps of
dark matter halos. \emph{Self-interacting} dark matter instead can form
cores but also the cores can collapse into a density peak.

>>>>

Dwarf galaxies host, in many ways, the most extreme galactic
environments in the universe. These little galaxies are typically
defined to be fainter than the LMC \citep[\(M_V \gtrsim -18\) or
similarly \(L \lesssim 10^9 L_\odot\),
e.g.,][]{hodge1971, mcconnachie2012}. As the smallest class of galaxies,
dwarfs are the most cosmologically numerous galaxies. Dwarf galaxies are
highly \emph{dark-matter dominated}, with mass to light ratios exceeding
100 to 1000 \(M_\odot/ L_\odot\). Because of their small total masses,
many dwarf galaxies are \emph{quenched}, with little to no recent star
formation. Their stellar populations are \emph{relics} from the early
universe, consisting of many of the oldest and most metal poor stars.
Understanding the properties of dwarf galaxies thus has implications
across astronomy, from cosmological structure formation on the smallest
scale to extremely metal-poor stellar populations to chemical evolution
to galactic dynamics. Of the classical dwarfs, the Sculptor and Ursa
Minor dwarf spheroidal galaxies stand out with a more extended density
profile relative to an exponential, hinting at tidal effects. As a case
study in tidal effects and our understanding of dwarf galaxy formation
and evolution, we aim to understand the dynamical history of these
galaxies.

@@

We only understand a tiny fraction of the universe's composition. The
leading theory of cosmology, \(\Lambda\)CDM (cold dark matter), posits
that the universe is composed of about 68\% dark energy (\(\Lambda\)),
27\% dark matter (DM), and 5\% regular baryons\footnote{Astronomers like
  to change definitions of words. \emph{Baryons} here means
  baryons+leptons, i.e.~any standard model massive fermion. The photon
  energy density is negligible today} \citep{planckcollaboration+2020}.
While the composition of dark matter and dark energy remains elusive, we
know their general properties. Dark energy causes the acceleration of
the expansion of the universe on large scales. We do not discuss dark
energy here---it does not substantially affect the local group today.
Dark matter, instead, makes up the vast majority of mass in galaxies.
Typically, galaxies have baryonic to dark matter ratios of between 1:5
to beyond 1:1000 for faint dwarf galaxies. In \(\Lambda\)CMD, dark
matter is assumed to only interact gravitationally, passing through
matter without effect (transparent to light, or \emph{dark}). Dark
matter is also \emph{cold}, i.e.~typical velocities much smaller than
the speed of light in the early universe. Implications of dark matter
properties range from cosmological structural formation, galaxy
structure, and galaxy interactions.

@@

Dark matter halos, and their associated galaxies, rarely evolve in
isolation. Instead, structure formation is \emph{hierarchical}. Small
dark matter halos collapse first and hierarchically merge into
progressively larger halos
\citep[e.g.,][]{blumenthal+1984, white+rees1978, white+frenk1991}.
Structure formation happens on a wide range of scales. The largest
structures become the cosmic web---composed of voids, filaments, and
clusters---and the smallest structures directly observable host dwarf
galaxies. Hierarchical assembly is evident through the large scale
structure of the universe, remnants of past mergers within the milky
way, and tidal disruption of dwarf galaxies and their streams around
nearby galaxies.

Small-scale structure formation is sensitive to deviations from
\(\Lambda\)CDM cosmology \citep[e.g.,][]{bechtol+2022}. Many alternative
models, such as warm dark matter or self-interacting dark matter, may
smooth out small-scale features and reduce the abundance of small halos
or change their structure \citep[e.g.,][]{lovell+2014}. Dwarf galaxies,
occupying the smallest dark matter halos, are promising windows into
small-scale cosmological features. Alternative dark matter properties
can influence dwarf galaxy abundance, formation, structure,, and tidal
evolution. The nearby dwarf galaxies of the local group, with detailed
observations, present a promising opportunity to understand the
evolution of these objects and test our understanding of cosmology. To
understand the evolution of these objects, we need to understand the
general predictions from \(\Lambda\)CDM for the properties of dwarf
galaxies.

@@

The total mass of an NFW profile diverges, so halos are commonly
characterized by an overdensity criterion. The virial mass, \(M_{200}\),
is defined as the mass within a region \(r_{200}\) containing a mean
enclosed density 200 times\footnote{From approximate analytic arguments
  about the mean density of a self-gravitating spherical collapse halo
  in an expanding universe, the expected virial overdensity is \ldots,
  commonly rounded to 200 (REFS).} the critical density of the universe:
\begin{equation}{
M_{200} =(4\pi/3) \ r_{200}^3\ 200\rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z) / 8\pi G
}\end{equation} A standard second parameter is the halo concentration,
\(c=r_{200} / r_s\) describing how the characteristic size of the halo
compares to its virial radius. In this case, the scale density is a
function of \(c\) alone,
\(\rho_s = (200/3)\,\rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\)
\citep{NFW1996}. However, we characterize halos by their circular
velocity profiles. The circular velocity,
\(v_{\rm circ}(r) = \sqrt{G M(r) / r}\), reaches a maximum of
\(v_{\rm max}\) at radius \(r_{\rm max} = \alpha\,r_s\) where
\(\alpha\approx2.16258\). \(v_{\rm max}\) is related to both the total
halo mass and the observed line of sight velocity disperion, and
\(r_{\rm max}\) relates to the scale radius.

While an NFW profile has two free parameters (\(M_{200}\) and \(c\)),
these are not independent. Smaller dark matter halos often collapse
earlier, when the universe was denser. As a result, small subhalos tend
to be more concentrated \citep[e.g.,][]{NFW1997}. The relationship
between \(M_{200}\) and c, or the mass-concentration relation, how more
massive halos become less concentrated in a predictable way
\citep[e.g.,][]{bullock+2001, ludlow+2016}. The lower panel of
Fig.~\ref{fig:smhm} illustrates the present-day mass-concentration
relation in terms of \(r_{\rm max}\) and \(v_{\rm max}\). While there is
a general trend in concentration with mass, the concentration values
still have substantial scatter. Other parameters such as the halo spin
or shape may add additional variability to present-day distribution of
halos, but this variation is typically expected to be small
\citep{navarro+2010, dicintio+2013, dutton+maccio2014}.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/example_density_profiles.pdf}}
\caption[Example density profiles]{Density profiles in log density
versus log radius for the stars and dark matter of a Sculptor-like
galaxy. The dark matter is more extended and massive than the star
across the entire galaxy---dynamical evolution is driven by the dark
matter mass distribution.}\label{fig:nfw_density}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/cosmological_means.pdf}}
\caption[Stellar-mass halo-mass relation]{\textbf{Top} Stellar mass
(top) as a function of maximum circular velocity. The solid line with
the 1-\(\sigma\) shaded region is the relation from \citet{fattahi+2018}
with scatter points simulated central galaxies from APOSTLE in
\citet{fattahi+2018}. \textbf{Bottom:} Radius of maximum circular
velocity \(r_{\rm max}\) as a function of maximum circular velocity. The
solid line with 1-\(\sigma\) shaded region and dashed line are the
relations from \citet{ludlow+2016} for \(z=0\) and \(z=2\) respectively.
\emph{TODO}: add legend / annotations.}\label{fig:smhm}
\end{figure}

@@

Both cosmology and observations find fundamental relationships between
the dynamical or total mass and stellar masses of galaxies. A key
prediction of \(\Lambda\)CDM is the stellar mass halo mass relation,
describing the amount of stellar mass forms in a (sub)halo of a given
size. In particular, the SMHM relation grows especially steep in the
dwarf galaxy regime---many dwarf galaxies are formed in halos of similar
masses. Fig.~\ref{fig:smhm} shows the stellar mass versus
\(v_{\rm max}\) maximum circular velocity (proxy for halo mass) for
apostle dwarfs from \citet{fattahi+2018}. While there is some scatter,
the range of predicted \(v_{\rm max}\) is fairly narrow across
\(\sim 4\) decades in stellar mass. Because lower mass galaxies have
increasingly shallow potential wells, feedback becomes more effective
removing gas more quickly. In addition, re-ionization suppresses late
star formation in the faintest galaxies. Thus, if we know the initial
stellar mass of a dwarf galaxy, we also know, at some level, the
properties of its dark matter halo.

Several challenges complicate a simple SMHM trend including environment,
assembly history, and tidal effects. Additionally, the detailed physics
of galaxy formation can impact the SMHM relationship. By being closer to
a massive host, many dwarfs quench earlier, resulting in lower stellar
masses \citep[e.g.,][]{christensen+2024}. Additionally, the time of
formation (relative to reionization) can influence the resulting stellar
content \citep{kim+2024}. Finally, tides influence both the dark matter
and stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
Consequently, tides may reduce the halo mass more than the stellar mass,
adding additional scatter to the SMHM trend, particularly for low-mass
satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
tides on local group dwarf galaxies is essential to determining where
and how these galaxies formed in a cosmological context.

@@

Dwarf galaxies have long raised conundrums in theories of galaxy
formation. The discovery of Fornax and Sculptor in 1938
\citep{shapley1938}\footnote{technically, the LMC and SMC may be
  classified as dwarf galaxies, but these were likely always known to
  humans at southern latitudes.}, with no known analogues, already
presented such an enigma. Shapley presented these dwarfs as a new type
of \emph{stellar system} resembling the Magellanic clouds and globular
clusters but did not attempt to speculate on the exact nature. While
dwarf galaxies were quickly understood to be galaxies based on the
inferred luminosities and sizes, their nature and formation remained
unclear for decades \citep[e.g.,][]{hodge1971, gallagher+wyse1994}.

The earliest spectroscopic work hinted that dwarf galaxies may contain
substantial dark matter masses. From early determinations of the
velocity dispersion for Sculptor and Ursa Minor dSph
\citep[e.g.,][\citet{aaronson+olszewski1987}]{aaronson1983}, inferred
mass-to-light ratios were at least 10 times the expectation for globular
clusters. While only determined by a few stars initially, these values
have solidified with time with larger and more precise samples
\citep[e.g.,][]{hargreaves+1994}. Subsequently, several theories
attempting to understand the formation and observed properties of these
objects were proposed. Examples include: dwarf galaxies are undergoing
tidal dissolution resulting in extreme mass-to-light ratios
\citep[e.g.,][]{kuhn+miller1989}, presence of massive central black
holes \citep[e.g.,][]{strobel+lake1994}, the formation of dark matter
free out-of-equilibrium ``tidal dwarfs'' from past mergers
\citep[e.g.,][\citet{kroupa1997}]{lynden-bell1982}, or modified theories
of gravity \citep{milgrom1995}. However, consistency with CDM galaxy
formation was also not out of the question
\citep[e.g.,][]{dekel+silk1986}. Since then, we have known that dwarf
galaxies are among the darkest objects in the universe, and
understanding their properties is critical to understanding the
universe.

Dwarf galaxies span a large range of physical sizes, luminosities, and
morphologies. Broadly, there are three classes of dwarf galaxies based
on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
\textbf{bright dwarfs galaxies} with magnitudes
\(-18 \lesssim M_V \lesssim  -14\), often exhibit irregular morphologies
and recent star formation. Fig.~\ref{fig:galaxy_images} shows the LMC as
an example of an irregular (and slightly bright) dwarf galaxy displaying
a bar. \textbf{Classical dwarf} galaxies occupy intermediate
luminosities ( \(-14 \lesssim M_V  \lesssim -7.7\)). Typically, these
systems are old, gas-poor, and spheroidal. All Milky Way satellites
discovered before digital sky surveys are classicals. The 12 classical
dwarfs satellites of our galaxy are Sagittarius, Fornax, Leo I,
Sculptor, Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici
I, Sextans I, and Crater II.\footnote{While formally the dwarf galaxy
  names we discuss contain'' dwarf spheroidal'' (dSph), e.g.~Sculptor
  dSph, we omit this suffix for brevity.} The \textbf{ultra-faint}s
occupy the very faintest magnitudes (\(-7.7 \lesssim M_V\)). These
galaxies have minuscule stellar masses, tend to be more compact, and are
the darkest known galaxies. Altogether, dwarf galaxies span about 7
orders of magnitude in stellar mass.

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.pdf}
\caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Fornax (DES
DR2), Sculptor (DES DR2), and Ursa Minor (UNWISE with Gaia point sources
over-plotted). Each image includes the galaxies luminosity and a 200 pc
scalebar. While the LMC is very prominent in the sky, the classical
dwarf galaxies are not obvious except in terms of star counts
(e.g.,Fig.~\ref{fig:scl_selection}). All (non-LMC) dwarfs occupy the
central \(\sim\)third of the image.}\label{fig:galaxy_images}
\end{figure}

@@

\emph{Gaia} has redefined astrometry, providing photometry, positions,
proper motions, and parallaxes for over 1 billion stars
\citep{gaialcollaboration+2021}. Launched in 2013, \emph{Gaia} is a
space-based, all-sky survey telescope situated at the Sun-Earth L2
Lagrange point \citep{gaiacollaboration+2016}. While \emph{Gaia}
completed its space-based mission in 2025, two forthcoming data releases
remain. Determining absolute parallax measurement is facilitated by the
observation that stars in different regions of the sky are affected by
parallax motion with different phases. By imagining two regions
separated by 106.5 degrees on the same focal plane, \emph{Gaia} measures
tiny changes in relative positions of stars across small and large
angles. Combining measurements from multiple epochs across several
years, an absolute all-sky reference frame is derived from which
parallax and proper motions are derived \citep{gaiacollaboration+2016}.
In addition to astrometry, \emph{Gaia} measures photometry in the wide
\emph{G} band (330-1050nm) and colours from the blue photometer (BP,
330-680 nm) and red photometer (RP, 640-1050 nm). \emph{Gaia}
additionally provides low resolution BP-RP spectra and radial velocity
measurements of bright stars (\emph{Gaia} radial velocity magnitudes
\textless16) \citep{gaiacollaboration+2016}. For this work,
\emph{Gaia}'s most relevant measurement are \(G\) magnitude,
\(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{The proper
  motions \(\mu_\alpha\) and \(\mu_\delta\) are the apparent rates of
  change in right ascension \(\alpha\) and declination \(\delta\),
  typically in units of mili-arcsecond (mas) per year. \(\mu_{\alpha*}\)
  allows where \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for
  projection effects in \(\alpha\).}

\emph{Gaia} has revolutionized many fields of astronomy including the
Milky Way and the Local Group. For example, 6D kinematic classification
of Milky Way stars has lead to the discovery of past mergers or Milky
Way building blocks like \emph{Gaia}-sausage Enceladus
\citep[e.g.,][]{helmi+2018}, out-of-equilibrium structures like the
\emph{Gaia} snail {[}e.g., \citet{antoja+2018l}, and dynamical effects
of the Milky Way's spiral arms and the bar in the solar neighbourhood
\citep{hunt+vasiliev2025}. Moving to the Milky Way halo, \emph{Gaia} has
helped find and constrain numerous stellar streams
\citep{bonaca+price-whelan2025}. Altogether, \emph{Gaia} has revealed
the hierarchical formation and complex, evolving structure of our own
galaxy.

For Milky Way satellites, \emph{Gaia} has enabled well-constrained
orbital analysis and facilitated precise stellar membership
determination. Before, proper motions of dwarf galaxies were spase or
undetermined. Few galaxies had precisely measured proper motions, often
from Hubble Space Telescope observations \citep[e.g.,][]{sohn+2017}.
\emph{Gaia} allowed for the first systematic determinations of Milky Way
satellite proper motions, and of unprecedented precision
\citep{MV2020a, pace+li2019}. While the proper motion uncertainty on a
typical dwarf member star is often with large uncertainties, by
combining the proper motions for 100s to 1000s of stars in \emph{Gaia},
incredibly precise proper motion measurements can be determined,
sometimes only limited by \emph{Gaia}'s systematic error floor
\citep[e.g.,][]{MV2020a}. Proper motions have ushered in a new dynamical
era for MW satellites studies, where we often have precise
determinations of the dwarf galaxy's recent orbit under a given MW
potential. In addition, \emph{Gaia} helps seperate out contaiminating MW
foreground stars. By measuring parallaxes or proper motions
significantly different from the dwarf galaxy, many stars can be
immediately removed as non-members
\citep[e.g.,][\citet{jensen2024}]{battaglia+2022}.

@@

Of local dwarfs, the classical systems still remain the best studied and
with most strongly constrained parameters. While extending 1-2 degrees
across the sky, making deriving properties trickier
\citep[e.g.,][]{mateo1998}, these dwarfs contain large numbers of bright
(giant) stars, allowing thousands of stars to be observed with deep
photometry and spectroscopy \citep[e.g.,][]{tolstoy+2023, pace+2020}. As
a result, the determination of fundamental properties such as the
position, size, orientation, distances, proper motions, line-of-sight
(LOS) velocities, and velocity dispersions \(\sigma_v\), are all
relatively well constrained and converged among different studies today.
However, ongoing research continues to redefine our understanding of the
detailed structure of Milky Way satellites, hinting that these objects
may be more complex than at first glance.

Observations of dwarf galaxies have been the origin of several disputes
or \emph{small-scale} problems in cosmology \citep[see
review][]{bullock+boylan-kolchin2017}. For example, the mismatch between
the expected number of dwarf galaxies from simulations and the observed
abundance was known as the \emph{missing satellites problem}.
Additionally, a number of observations showed that many dwarf galaxies,
but not all, possess dark matter ``cores'' as measured from circular
velocity curves \citep[e.g.,][]{moore1994, adams+2014, oh+2015} or dwarf
spheroidal \citep[albeit less constraining,
e.g.,][]{walker+penarrubia2011, read+walker+steger2019}. As a result,
numerous alternative forms of dark matter have been advocated as
solutions. \emph{Warm} dark matter, relativistic in the early universe
but cooler now, smooths out small-scale features and softens the cusps
of dark matter halos. \emph{Self-interacting} dark matter instead can
form cores but also the cores can collapse into a density peak.
